Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves

### 2025-10-19 ms/ms.tex
updated ms with new analysis

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gias}{$g_\text{ias}$}
\newcommand{\gmax}{$g_\text{max}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\usepackage{setspace}
\usepackage[modulo]{lineno}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves},
  pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Plasticity and adaptation to high light intensity amplify the
advantage of amphistomatous leaves}
\author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
Lim\(^{1}\) \and Dachuan Wang\(^{1,2}\)}
\date{}

\begin{document}
\maketitle

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were an
essential innovation in vascular plants, permitting them to grow tall on
land by enabling access to CO\(_2\) for photosynthetic carbon
assimilation while preventing hydraulic failure in variable environments
(1--3). Optimal stomatal function depends on both dynamic changes in
aperture on the scale of minutes to hours, as well as static anatomy
determined by developmental plasticity and constitutive genetic
differences (4--6). Understanding how stomata respond to environmental
change over daily, developmental, and evolutionary time is important for
studying adaptation (1, 7, 8), inferring paleoclimate from fossil
cuticles (9--11), predicting responses to climate change (12--15), and
improving crops (16). Stomatal function contributes to global carbon and
water cycles (17) and therefore to predicting future climate (18).

Despite extensive theoretical and empirical progress understanding
stomata function and anatomy from molecular to ecosystem levels, the
adaptive significance of amphistomatous leaves remains an important
unsolved problem in leaf structure-function relationships (19--26).
Amphistomatous leaves develop abaxial and adaxial stomata whose aperture
can be independently regulated (27--31) to control gas exchange through
each surface. All else being equal, simultaneous gas exchange through
stomata on both surfaces increases CO\(_2\) supply to chloroplasts by
providing a second parallel pathway through leaf intercellular
airspaces, enhancing photosynthesis (20, 32). The extent to which
amphistomy increases CO\(_2\) supply depends on resistance to diffusion
in intercellular airspaces. This resistance can be low in thin, porous,
amphistomatous leaves (28, 33), but may be more substantial in thick,
dense, hypostomatous leaves (34). We refer to the intercellular airspace
conductance (\gias), the inverse of resistance. Amphistomatous leaves
also lose more water through evaporation because of a second boundary
layer conductance (35), but the additional carbon gain should be enough
to offset this cost in most realistic scenarios (36).

@@

\includegraphics{../figures/hypotheses.pdf}

@@

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
(50)) from different native light habitats, grown under simulated sun
and shade light treatments, and measured under contrasting light
intensities (low and high). We measured \aax{} on 572 individual plants
from 29 populations (average of 9.86 replicates per light treatment)
using a recently developed method (46). With this method, we directly
compare the photosynthetic rate of an untreated amphistomatous leaf to
that of the same leaf with gas exchange blocked through the adaxial
(upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf stomatal conductance (\gsw), we measure \(A\) over
a range of \gsw, inducing stomatal opening and closure by modulating
humidity (see supplementary text for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (20), but with modifications
previously described in (46) and here (Supplementary Materials). The
native light intensity was represented by plant area index (PAI
\(\unit{\meter\squared\per\meter\squared}\)), estimated using a global
gridded data set derived from the Global Ecosystem Dynamics
Investigation {[}GEDI; (51){]} and georeferenced accession collection
information from the Tomato Genetics Resource Center
(Table~\ref{tbl-populations}). The growth light intensities were
\ppfdequals{761} (sun treatment) and \ppfdqty{115} (shade treatment)
while all other environment conditions were nearly identical (see
Materials and Methods). The high and low measurement light intensities
were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
(87.0:13.0 red:blue), respectively.

Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). Bayesian
phylogenetic mixed effects models that allowed \aax{} to vary between
measurement light intensities, growth light intensities, and among
populations outperformed simpler models based on information criteria
(Table~\ref{tbl-aa_loo1}). Measured under high light intensity, \aax{}
was consistently greater for sun plants. The average \aax{} among
populations in the shade treatment was 0.041 (range: 0.007--0.113; 19 of
29 populations significant); however, the same populations grown at high
light intensity showed a mean \aax{} of 0.052 (range: 0.020--0.120; 20
of 29 populations significant). Contrary to the predictions of the
acclimation hypothesis, \aax{} was greater in all populations under low
measurement light intensity for both sun and shade grown plants. The
overall average \aax{} of shade and sun grown plants measured under low
light intensity was 0.064 (range: 0.022--0.137; 28 of 29 populations
significant) and 0.100 (range: 0.049--0.206; 27 of 29 populations
significant), respectively. There was a modest tendency for populations
from more open habitats (lower PAI) to exhibit greater \aax{} and the
slope was significantly different than 0 in 3 of the 4 treatment
combinations (Fig.~\ref{fig-aa}B).

@@

\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) is greater in wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.}
\aax{} (\(y\)-axis) is the log-response ratio of photosynthesis in an
amphistomatous leaf compared to an otherwise identical
pseudohypostomatous leaf. In both panels, estimates are shown for plants
grown under simulated shade (brown) or sun (orange) and measured under
high light intensity (\ppfdequals{2000}; open circles) and low light
intensity (\ppfdequals{150}; solid circles). The dashed line at zero
indicates no difference in photosynthesis between amphi- and
pseudohypostomatous leaves. (\textbf{A}) The points are the posterior
median \aax{} for each population, with error bars showing 95\%
confidence intervals. Within each facet, the populations are arranged by
\aax{} estimated in that growth and measurement condition. The average
\aax{} among populations for each treatment group is written in the top
left corner. (\textbf{B}) The same estimates of \aax{} in each
combination of growth and measurement light intensity plotted against
native plant area index (PAI; \(x\)-axis, log-scale). The confidence
intervals were omitted for visual clarity. The median relationship
between native PAI (\(x\)-axis; log-scale) and \aax{} is shown as a
solid line with the shaded region between dashed lines showing the 95\%
confidence ribbon. The * in the top left corner of each facet indicates
the slope of the linear regression is significantly different from zero;
n.s. indicates not significant.}

@@

\bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
\CSLLeftMargin{60. }%
\CSLRightInline{Stan Development Team,
\emph{\href{https://mc-stan.org}{Stan {Modeling} {Language} {Users}
{Guide} and {Reference} {Manual}}} (2025).}

\bibitem[\citeproctext]{ref-burkner_brms_2017}
\CSLLeftMargin{61. }%
\CSLRightInline{P.-C. Bürkner,
\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
(2017).}

\bibitem[\citeproctext]{ref-gabry_cmdstanr_2025}
\CSLLeftMargin{62. }%
\CSLRightInline{J. Gabry, R. Češnovar, A. Johnson, S. Bronder,
\emph{\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{Cmdstanr:
{R} {Interface} to '{CmdStan}'}} (2025).}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{hyperref}
\addtokomafont{disposition}{\rmfamily}
\usepackage{siunitx}
\newcommand{\aahigh}{\mathrm{AA}_{2000}}
\newcommand{\aalow}{\mathrm{AA}_{150}}
\newcommand{\aashade}{\mathrm{AA}_{\text{shade}}}
\newcommand{\aasun}{\mathrm{AA}_{\text{sun}}}
\newcommand{\aax}{$\mathrm{AA}$}
\newcommand{\aay}{\mathrm{AA}}
\newcommand{\Aamphi}{$A_{\mathrm{amphi}}$}
\newcommand{\Ahypo}{$A_{\mathrm{hypo}}$}
\newcommand{\agcurve}{$A \textendash g_\text{sw}$}
\newcommand{\aqcurve}{$A \textendash Q$}
\newcommand{\ca}{$C_\mathrm{a}$}
\newcommand{\caequals}[1]{$C_\mathrm{a} = \qty{#1}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\cabetween}[2]{#1 < $C_\mathrm{a} < \qty{#2}{\micro\mol\raiseto{-1}\mol}$}
\newcommand{\gias}{$g_\text{ias}$}
\newcommand{\gmax}{$g_\text{max}$}
\newcommand{\gmaxratio}{$g_\text{max,ratio}$}
\newcommand{\gmaxshade}{$g_\text{max,shade}$}
\newcommand{\gmaxsun}{$g_\text{max,sun}$}
\newcommand{\gsw}{$g_\text{sw}$}
\newcommand{\loggsw}{$\log(g_\text{sw})$}
\newcommand{\logA}{$\log(A)$}
\newcommand{\ppfd}{$\mathrm{PPFD}$}
\newcommand{\ppfdequals}[1]{$\mathrm{PPFD} = \qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\ppfdqty}[1]{$\qty{#1}{\micro\mol\raiseto{-2}\meter\raiseto{-1}\second}$}
\newcommand{\rh}{$\mathrm{RH}$}
\newcommand{\rhequals}[1]{$\mathrm{RH} = #1\%$}
\newcommand{\tleafequals}[1]{$T_\mathrm{leaf} = \qty{#1}{\degreeCelsius}$}
\usepackage{setspace}
\usepackage[modulo]{lineno}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Plasticity and adaptation to high light intensity amplify the advantage of amphistomatous leaves},
  pdfauthor={Christopher D. Muir\^{}\{1,2*\}; Wei Shen Lim\^{}\{1\}; Dachuan Wang\^{}\{1,2\}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Plasticity and adaptation to high light intensity amplify the
advantage of amphistomatous leaves}
\author{Christopher D. Muir\(^{1,2*}\) \and Wei Shen
Lim\(^{1}\) \and Dachuan Wang\(^{1,2}\)}
\date{}
\begin{document}
\maketitle

@@

Stomata are microscopic pores on the surfaces of leaves and other
photosynthetic organs formed by a pair of guard cells. They are
essential for balancing carbon gained per unit water lost and were a key
innovation in vascular plants, permitting them to grow tall on land by
enabling access to CO\(_2\) for photosynthetic carbon assimilation while
preventing hydraulic failure in variable environments (1--3). Optimal
stomatal function depends on both dynamic changes in aperture on the
scale of minutes to hours, as well as static anatomy determined by
developmental plasticity and constitutive genetic differences (4--6).
Understanding how stomata respond to environmental change over daily,
developmental, and evolutionary time is important for studying
adaptation (1, 7, 8), inferring paleoclimate from fossil cuticles
(9--11), predicting responses to climate change (12--15), and improving
crops (16). Stomatal function contributes to global carbon and water
cycles (17) and therefore to predicting future climate (18).

Despite extensive theoretical and empirical progress understanding
stomatal function and anatomy from molecular to ecosystem levels, the
adaptive significance of amphistomatous leaves remains an important
unsolved problem in leaf structure-function relationships (19--26).
Amphistomatous leaves develop abaxial and adaxial stomata whose aperture
can be independently regulated (27--31) to control gas exchange through
each surface. All else being equal, simultaneous gas exchange through
stomata on both surfaces increases CO\(_2\) supply to chloroplasts by
providing a second parallel pathway through leaf intercellular
airspaces, enhancing photosynthesis (20, 32). The extent to which
amphistomy increases CO\(_2\) supply depends on resistance to diffusion
in intercellular airspaces. This resistance can be low in thin, porous,
amphistomatous leaves (28, 33), but may be more substantial in thick,
dense, hypostomatous leaves (34). We refer to the intercellular airspace
conductance (\gias), the inverse of resistance. Amphistomatous leaves
also lose more water through evaporation because of a second boundary
layer conductance (35), but the additional carbon gain should be enough
to offset this cost in most realistic scenarios (36).

@@

\pandocbounded{\includegraphics[keepaspectratio]{../figures/hypotheses.pdf}}

@@

We tested these hypotheses by comparing \aax{} among amphistomatous wild
tomato species (Table~\ref{tbl-populations}; Figure~\ref{fig-sr-rxnorm};
(50)) from different native light habitats, grown under simulated sun
and shade light treatments, and measured under contrasting light
intensities (low and high). We measured \aax{} on 567 individual plants
from 29 populations (average of 9.78 replicates per light treatment)
using a recently developed method (46). With this method, we directly
compare the photosynthetic rate of an untreated amphistomatous leaf to
that of the same leaf with gas exchange blocked through the adaxial
(upper) surface by transparent plastic, which we refer to as
`pseudohypostomy'. To compare amphi- and pseudohypostomatous leaves at
identical whole-leaf stomatal conductance (\gsw), we measure \(A\) over
a range of \gsw, inducing stomatal opening and closure by modulating
humidity (see supplementary text for further details). We estimated
`amphistomy advantage' (\aax) \emph{sensu} (20), but with modifications
previously described in (46) and here (Supplementary Materials). The
native light intensity was represented by plant area index (PAI
\(\unit{\meter\squared\per\meter\squared}\)), estimated using a global
gridded data set derived from the Global Ecosystem Dynamics
Investigation {[}GEDI; (51){]} and georeferenced accession collection
information from the Tomato Genetics Resource Center
(Table~\ref{tbl-populations}). The growth light intensities were
\ppfdequals{761} (sun treatment) and \ppfdqty{115} (shade treatment)
while all other environment conditions were nearly identical (see
Materials and Methods). The high and low measurement light intensities
were \ppfdequals{2000} (97.8:2.24 red:blue) and \ppfdequals{150}
(87.0:13.0 red:blue), respectively.

Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). We tested
whether \aax{} varied among treatments by comparing the fit of competing
Bayesian phylogenetic mixed effects models using information criteria
(see Materials and Methods). Models including measurement and growth
light intensity consistently outperformed simpler models
(Table~\ref{tbl-aa_loo1}). There was equivocal support for interactions
between factors and population because several models including these
terms performed similarly to those without them
(Table~\ref{tbl-aa_loo1}). We focus on results that were robust to model
structure and addressed our \emph{a priori} hypotheses using parameter
estimates from the best model.

Measured under high light intensity, \aax{} was consistently greater for
sun plants. The average \aax{} among populations in the shade treatment
was 0.039 (range: 0.012--0.098; 23 of 29 populations significant);
however, the same populations grown at high light intensity showed a
mean \aax{} of 0.055 (range: 0.028--0.114; 28 of 29 populations
significant). Contrary to the predictions of the acclimation hypothesis,
\aax{} was greater in all populations under low measurement light
intensity for both sun and shade grown plants. The overall average
\aax{} of shade and sun grown plants measured under low light intensity
was 0.074 (range: 0.032--0.141; 28 of 29 populations significant) and
0.090 (range: 0.049--0.158; 29 of 29 populations significant),
respectively. There was a modest tendency for populations from more open
habitats (lower PAI) to exhibit greater \aax{} and the slope was
significantly different than 0 in all treatment combinations
(Fig.~\ref{fig-aa}B).

@@

\caption{\label{fig-aacap}\textbf{Amphistomy advantage (\aax{}) is greater in wild tomatoes from open habitats and is further amplified by developmental plasticity under sunny growth conditions.}
\aax{} (\(y\)-axis) is the log-response ratio of photosynthesis in an
amphistomatous leaf compared to an otherwise identical
pseudohypostomatous leaf. In both panels, estimates are shown for plants
grown under simulated shade (brown) or sun (orange) and measured under
high light intensity (\ppfdequals{2000}; open circles) and low light
intensity (\ppfdequals{150}; solid circles). The dashed line at zero
indicates no difference in photosynthesis between amphi- and
pseudohypostomatous leaves. (\textbf{A}) The points are the posterior
median \aax{} for each population, with error bars showing 95\%
confidence intervals. Within each facet, the populations are arranged by
\aax{} estimated in that growth and measurement condition. The average
\aax{} among populations for each treatment group is written in the top
left corner. (\textbf{B}) The same estimates of \aax{} in each
combination of growth and measurement light intensity plotted against
native plant area index (PAI; \(x\)-axis, log-scale). The confidence
intervals were omitted for visual clarity. The median relationship
between native PAI (\(x\)-axis; log-scale) and \aax{} is shown as a
solid line with the shaded region between dashed lines showing the 95\%
confidence ribbon. The * in the top left corner of each facet indicates
the slope of the linear regression is significantly different from
zero.}

@@

\bibitem[\citeproctext]{ref-saathoff_gas_2021}
\CSLLeftMargin{60. }%
\CSLRightInline{A. J. Saathoff, J. Welles,
\href{https://doi.org/10.1111/pce.14178}{Gas exchange measurements in
the unsteady state}. \emph{Plant, Cell \& Environment} \textbf{44},
3509--3523 (2021).}

\bibitem[\citeproctext]{ref-stan_development_team_stan_2025}
\CSLLeftMargin{61. }%
\CSLRightInline{Stan Development Team,
\emph{\href{https://mc-stan.org}{Stan {Modeling} {Language} {Users}
{Guide} and {Reference} {Manual}}} (2025).}

\bibitem[\citeproctext]{ref-burkner_brms_2017}
\CSLLeftMargin{62. }%
\CSLRightInline{P.-C. Bürkner,
\href{https://doi.org/10.18637/jss.v080.i01}{\textbf{Brms} : {An}
\emph{r} {Package} for {Bayesian} {Multilevel} {Models} {Using}
\emph{stan}}. \emph{Journal of Statistical Software} \textbf{80}
(2017).}
